BOLD Parameter Estimation Using Particle Filters

### 2010-08-14 doc/thesis/results.tex
fixes to results

\section{Single Time Series}
\label{sec:Single Voxel Simulation}
Given the state-space equations for the BOLD signal, simulating a single time 
series is relatively straight forward. After generating a true signal,
identically and independently distributed (I.I.D.) Gaussian noise and a Wiener
process with Gaussian I.I.D. steps wered added to this true signal. Finally a 
carrier level is added, since BOLD is typically
measured as a \% difference from the baseline. The particle filter
algorithm will immediately remove this by calculating the \% difference, 
but adding a carrier level meant that the exact same algorithm used 
for simulated data could be used for the real data.

Once this noisy simulated time series was generated, the particle filter algorithm
 was then run on single time series image. Here I will include two sets of tests 
to determine the power of the particle filter in modeling. The first set are 
used to demonstrate the ability of the particle filter to find the most likely
set of parameters/state variables over the course of the run. The second type
of test will investigate the problem of false-positives. As the first
round of tests will show, given that there is an underlying BOLD signal, 
the particle filter is excellent at finding the most probable cause of 
the observed signal. However, as discussed in the false positive tests,
even if there is no underlying signal, the particle filter will still determine
the most likely BOLD signal; which may not be flat. Thus, it is 
important to investigate methods of determining false positives using 
other methods.

%LOW NOISE SECTION, with a signal
\subsection{Signal with Low Noise}
\label{sec:SimLowNoise}
%tests had noise of $ \{\sigma_y = .01, \sigma_x = .005\} $, where $\sigma_y$ is the
%measurement noise, and $\sigma_x$ is the wiener step size. Both signals used the
%parameters .
%The particle filter used the parameters defined in \autoref{tab:Prior} (\autoref{sec:PriorDistrib}),
%thus the particle filter was not centered over the correct values. 
\begin{figure}
\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=7cm]{images/realization_lownoise}
\caption{Test Signals with low noise compared to the clean signal.}
\label{fig:LowNoiseRealization}
\end{figure}
\begin{figure}
\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=7cm]{images/preprocessed_lownoise}
\caption{A comparison of the preprocessed signals for the low noise case. This is the
noisy input to the actual particle filter algorithm.}
\label{fig:PreprocessedLowNoise}
\end{figure}
\begin{figure}
\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/comparison_lownoise}
\caption{A comparison of the fitted signals for the low noise case.}
\label{fig:FitComparisonLowNoise}
\end{figure}

To begin the single voxel simulation; I generated a signal using the following parameters:
$\{\tau_0 = 1.45, \alpha = .3, E_0 = .47, V_0 = .044, \tau_s = 1.94, \tau_f = 1.99, \epsilon = 1.8\}$.
These same parameters were also used in the high noise case. The noise realizations were
then generated based on a measurement noise ($\sigma_y$) of $.001$ and a drift standard deviation
($\sigma_x$) of $.0005$. The measurement noise as well as the steps of the drift
were taken to be Gaussian
For this low noise case, the eleven realizations are shown in \autoref{fig:LowNoiseRealization}.
The actual signal delivered into the particle filter
was the result of some preprocessing to remove drift, as described in 
\autoref{sec:Methods Preprocessing}. The bias introduced into the signal by preprocessing 
may have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
The preprocessing consists of several steps, which are discussed in detail in \autoref{sec:Methods Preprocessing}.
The data shows that the spline struggles to match the signal near the end, although 
overall, the preprocessing seems to have been successful at removing trends. Finally, the 
set of fits to the true BOLD signal are shown in \autoref{fig:FitComparisonLowNoise}.
The filtering property of the particle filter is clear from the results here.  In several locations the output
of the particle filter looks like a filtered version of the input. For instance toward the
end, the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm has converged sufficiently to know that that such a movement in
signal isn't possible. A similar circumstance occurs at 100 seconds in. 
It is also worth noting the bias introduced by pre-processing 
results in a slightly larger peak signal. The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}. Note that throughout the results, unless otherwise specified,
the residual will be defined as the square root of the mean squared residual, 
\begin{equation}
\text{Residual} = \sqrt{\sum (\hat{y}_k - y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output at time $k$ and $y_k$ is the output
sample at time $k$. Note the subtle distinction between this and the error
which is defines as 
\begin{equation}
\text{Error} = \sqrt{\sum (\hat{y}_k - Y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output, and $Y_k$ is the underlying (free of noise)
output. Throughout this section and the next, the terms \emph{error} and \emph{residual}
will be referencing these values, for an entire timeseries.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c | c  c  c  c  c  c  c |}
\hline 
  & $\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$ \\
\hline 
\rowcolor[gray]{.8} $\tau_0$ &   0.0189481 & -0.0014269 & -0.0011267 & -1.13e-05 & -0.0025616 & -0.0189559 & 0.0070405 \\
$\alpha$ &                       -0.0014269 & 0.0026716 & 9.93e-05 & -0.0002041 & -0.0008632 & -0.0016823 & 0.0071891 \\
\rowcolor[gray]{.8} $E_0$    &   -0.0011267 & 9.93e-05 & 0.0010701 & -0.0002277 & -0.0001177 & 0.0001013 & 0.0016972 \\
$V_0$    &                       -1.13e-05 & -0.0002041 & -0.0002277 & 0.0005401 & 4.3e-06 & 4.56e-05 & -0.0080494 \\
\rowcolor[gray]{.8} $\tau_s$ &   -0.0025616 & -0.0008632 & -0.0001177 & 4.3e-06 & 0.0128056 & 0.012878 & -0.005516 \\
$\tau_f$ &                       -0.0189559 & -0.0016823 & 0.0001013 & 4.56e-05 & 0.012878 & 0.0416927 & -0.0158182 \\
\rowcolor[gray]{.8} $\epsilon$&  0.0070405 & 0.0071891 & 0.0016972 & -0.0080494 & -0.005516 & -0.0158182 & 0.1567165 \\
\hline 
\end{tabular}
\caption{Typical Covariance matrix of the parameters at the end of a run.}
\label{tab:CovSim} 
\end{table}

@@

There are a few results worth noting in the final parameter estimates (which 
are the mean of the particle filter's posterior distribution). First the time constants vary 
greatly across
runs with different noise realizations, yet the sum of the individual time constants
($\tau_f$, $\tau_s$ and $\tau_0$) seems to be more consistent. In general the 
time constants fall short of the true time constant. This could be a limitation
based on the prior distribution (which notably has an initial mean below the true values) 
or it could be caused by the disassociated benefits of a correct time
constant with regards to a correct output. It often takes several measurement periods before 
a difference
in time constants becomes relevant to the weight. Its also possible that output is insensitive
to differences in the time constants; althouhg varying the activation durations could induce
better estimates.  The large variation of $V_0$ are also interesting, although as one final
covariance matrix shows, there is a definite negative correlation between $\epsilon$ and $V_0$.
In general,
with this admittedly small amount of noise, it would appear that the relation between a set
of parameters/stimuli and the output is not injective; a time
series is not unique to a single set of parameters. This is good justification that 
simultaneous blood volume or tagged flow calculations with the conventional FMRI 
could benefit the model.

The covariance matrix (\autoref{tab:CovSim}) provides a great deal of insight into 
the way parameters evolve in the particle filter. The fact that there are so many 
covariance elements that are as big or bigger than variance elements indicates just
loose the parameters set is. Every single parameters, including $\alpha$ seems to
have a non-zero covariance. Why $\alpha$ and $\epsilon$ would be positively related
is one of the most interesting element. Given the difficulty in mapping out a 
seven dimensional space, it is difficult to know why such correlations exist. 
 Notice the covariance of $\tau_f$ and $\tau_0$
is $-0.019$ whereas
the variance of $\tau_0$ and $\tau_f$ are $0.019$ and $0.04$ respectively. Clearly there is 
some measure of ill-defined behavior between the different time-constants, which is to
be expected given the number steps before a stimuli affects the output. 
The convergence properties of the first run in \autoref{tab:LowNoiseResults} 
demonstrates the migration of parameters to match the estimates. Given the relatively high
variance of parameters, its likely that more measurements or perhaps more variation
in the stimuli could further differentiate the parameters. Its also notable that
the final mean of the parameters may not be the best point estimator, and that perhaps
using the mode would be a better way to estimate the output. Regardless, the time series'
generated from the estimated mean seems to match the correct signal very well 
(\autoref{fig:FitComparisonLowNoise}).

@@

>>>>

\section{Single Time Series}
\label{sec:Single Voxel Simulation}
Given the state-space equations for the BOLD signal, simulating a single time 
series was relatively straight forward. After generating a true signal,
identically and independently distributed (I.I.D.) Gaussian noise and a Wiener
process with Gaussian I.I.D. steps were added to this true signal. Finally a 
carrier level was added, since BOLD is typically
measured as a \% difference from the baseline. The particle filter
algorithm will immediately remove this by calculating the \% difference, 
but adding a carrier level meant that the exact same algorithm used 
for simulated data could be used for the real data.

Once this noisy simulated time series was generated, the particle filter algorithm
 was run on the single time series image. Here I include four sets of tests 
to determine the power of the particle filter in modeling. The first two were
used to demonstrate the ability of the particle filter to find the most likely
set of parameters/state variables over the course of the run. The second two 
tests investigated the problem of false-positives. As the first
round of tests show, given that there is an underlying BOLD signal, 
the particle filter is excellent at finding the most probable cause of 
the observed signal. However, as discussed in the false positive tests,
even if when there was no underlying signal, the particle filter still determined
the most likely BOLD signal; which was not 0. Thus, it was
important to investigate methods of determining false positives using 
other methods.

%LOW NOISE SECTION, with a signal
\subsection{Signal with Low Noise}
\label{sec:SimLowNoise}
%tests had noise of $ \{\sigma_y = .01, \sigma_x = .005\} $, where $\sigma_y$ is the
%measurement noise, and $\sigma_x$ is the wiener step size. Both signals used the
%parameters .
%The particle filter used the parameters defined in \autoref{tab:Prior} (\autoref{sec:PriorDistrib}),
%thus the particle filter was not centered over the correct values. 
\begin{figure}
\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/realization_lownoise}
\caption{Test Signals with low noise compared to the clean signal.}
\label{fig:LowNoiseRealization}
\end{figure}
\begin{figure}
\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/preprocessed_lownoise}
\caption{A comparison of the preprocessed signals for the low noise case. This is the
noisy input to the actual particle filter algorithm.}
\label{fig:PreprocessedLowNoise}
\end{figure}
\begin{figure}
\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/comparison_lownoise}
\caption{A comparison of the fitted signals for the low noise case.}
\label{fig:FitComparisonLowNoise}
\end{figure}

To begin the single voxel simulation; I generated a signal using the following parameters:
$\{\tau_0 = 1.45, \alpha = .3, E_0 = .47, V_0 = .044, \tau_s = 1.94, \tau_f = 1.99, \epsilon = 1.8\}$.
These same parameters were also used in the high noise case. The noise realizations were
then generated based on a measurement noise ($\sigma_y$) of $.001$ and a drift standard deviation
($\sigma_x$) of $.0005$. The measurement noise as well as the steps of the drift
were taken to be Gaussian.
For this low noise case, the eleven realizations are shown in \autoref{fig:LowNoiseRealization}.
The actual signal delivered into the particle filter
was the result of preprocessing to remove drift, as described in 
\autoref{sec:Methods Preprocessing}. The bias introduced into the signal by preprocessing 
may have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
The data shows that the spline struggles to match the signal near the end, although 
overall, the preprocessing seems to have been successful at removing trends. While finding
a better preprocessing algorithm could be useful, non will ever be perfect. In this case
the cause of the problem was the final stimuli that pushed the final median well above the base
level. The fitting lines to the true BOLD signal are shown in \autoref{fig:FitComparisonLowNoise}.
 In several locations the output
of the particle filter looks like a filtered version of the input. For instance toward the
end, the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm has converged sufficiently to prevent inexplicable movement.
A similar circumstance occurs at 100 seconds. Note that bias introduced by pre-processing 
results in a slightly larger peak signal. The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}. Note that throughout the results, unless otherwise specified,
the term \emph{residual} will be defined as the square root of the mean squared residual, 
\begin{equation}
\text{Residual} = \sqrt{\sum (\hat{y}_k - y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output at time $k$ and $y_k$ is the output
sample at time $k$. Note the subtle distinction between this and what will be 
termed the \emph{error} which is defines as 
\begin{equation}
\text{Error} = \sqrt{\sum (\hat{y}_k - Y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output, and $Y_k$ is the underlying (free of noise)
signal. Throughout this section and the next, the terms error and residual
will be referencing these values (for a particular timeseries).

@@

%\begin{table}[t]
%\centering
%\begin{tabular}{|c | c  c  c  c  c  c  c |}
%\hline 
%  & $\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$ \\
%\hline 
%\rowcolor[gray]{.8} $\tau_0$ &   0.0189481 & -0.0014269 & -0.0011267 & -1.13e-05 & -0.0025616 & -0.0189559 & 0.0070405 \\
%$\alpha$ &                       -0.0014269 & 0.0026716 & 9.93e-05 & -0.0002041 & -0.0008632 & -0.0016823 & 0.0071891 \\
%\rowcolor[gray]{.8} $E_0$    &   -0.0011267 & 9.93e-05 & 0.0010701 & -0.0002277 & -0.0001177 & 0.0001013 & 0.0016972 \\
%$V_0$    &                       -1.13e-05 & -0.0002041 & -0.0002277 & 0.0005401 & 4.3e-06 & 4.56e-05 & -0.0080494 \\
%\rowcolor[gray]{.8} $\tau_s$ &   -0.0025616 & -0.0008632 & -0.0001177 & 4.3e-06 & 0.0128056 & 0.012878 & -0.005516 \\
%$\tau_f$ &                       -0.0189559 & -0.0016823 & 0.0001013 & 4.56e-05 & 0.012878 & 0.0416927 & -0.0158182 \\
%\rowcolor[gray]{.8} $\epsilon$&  0.0070405 & 0.0071891 & 0.0016972 & -0.0080494 & -0.005516 & -0.0158182 & 0.1567165 \\
%\hline 
%\end{tabular}
%\caption{Typical Covariance matrix of the parameters at the end of a run.}
%\label{tab:CovSim} 
%\end{table}

@@

There are a few important results in the final parameter estimates (which 
are the mean of the particle filter's posterior distribution). First, the time constants vary 
greatly across
runs with different noise realizations, yet the sum of the individual time constants
($\tau_f$, $\tau_s$ and $\tau_0$) seems to be more consistent. In general the 
time constants fall short of the true time constant. This could be a limitation
based on the prior distribution (which notably has an initial mean below the true values) 
or it could be other parameters are compensating. Its also possible that output is insensitive
to differences in the time constants; although varying the activation durations could induce
better estimates.  
The convergence properties of the first run in \autoref{tab:LowNoiseResults} 
demonstrates the migration of parameters to match the estimates. In spite of the
rather wide final parameter variances, the time series'
generated from the estimated mean seems to match the correct signal very well 
(\autoref{fig:FitComparisonLowNoise}).

%The large variation of $V_0$ are also interesting, although as the final
%covariance matrix of the first run shows, there is a significant correlation between $\epsilon$ and $V_0$.
%In general, with even a small amount of noise, parameters are ill defined. Thus at
%least for this set of stimuli the parameters are not uniquely identifiable. 
%
%The covariance matrix (\autoref{tab:CovSim}) provides a great deal of insight into 
%the way parameters evolve in the particle filter. The number of 
%covariance elements that exceed the variance elements indicates the
%significant interplay between parameters. Every single parameters, including $\alpha$, 
%has a non-zero covariance. Why $\alpha$ and $\epsilon$ would be positively related
%is the most interesting element. The answer of how exactly each parameter
%compensates for other parameters is complex.  Notice the covariance of $\tau_f$ and $\tau_0$
%is $-0.019$ whereas the variance of $\tau_0$ and $\tau_f$ are $0.019$ and $0.04$ respectively. 
%%stopped here
%As Clearly there is some measure of ill-defined behavior between the different time-constants, which is to
%be expected given the number steps before a stimuli affects the output. 
%The convergence properties of the first run in \autoref{tab:LowNoiseResults} 
%demonstrates the migration of parameters to match the estimates. Given the relatively high
%variance of parameters, its likely that more measurements or perhaps more variation
%in the stimuli could further differentiate the parameters. Its also notable that
%the final mean of the parameters may not be the best point estimator, and that perhaps
%using the mode would be a better way to estimate the output. Regardless, the time series'
%generated from the estimated mean seems to match the correct signal very well 
%(\autoref{fig:FitComparisonLowNoise}).

@@

\subsection{Very Long FMRI Run}
\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=16cm]{images/longfit}}
\caption{BOLD estimate converging for a very long FMRI run.}
\label{fig:long_preprocessed}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=16cm]{images/longfit_500}}
\caption{BOLD estimate converging for a very long FMRI run, first 500 seconds.}
\label{fig:long_preprocessed}
\end{figure}

\begin{table}[t]
\centering
\begin{tabular}{|c | c  c  c  c  c  c  c |}
\hline 
  & $\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$ \\
\hline 
\rowcolor[gray]{.8} $\tau_0$ &   0.0189481 & -0.0014269 & -0.0011267 & -1.13e-05 & -0.0025616 & -0.0189559 & 0.0070405 \\
$\alpha$ &                       -0.0014269 & 0.0026716 & 9.93e-05 & -0.0002041 & -0.0008632 & -0.0016823 & 0.0071891 \\
\rowcolor[gray]{.8} $E_0$    &   -0.0011267 & 9.93e-05 & 0.0010701 & -0.0002277 & -0.0001177 & 0.0001013 & 0.0016972 \\
$V_0$    &                       -1.13e-05 & -0.0002041 & -0.0002277 & 0.0005401 & 4.3e-06 & 4.56e-05 & -0.0080494 \\
\rowcolor[gray]{.8} $\tau_s$ &   -0.0025616 & -0.0008632 & -0.0001177 & 4.3e-06 & 0.0128056 & 0.012878 & -0.005516 \\
$\tau_f$ &                       -0.0189559 & -0.0016823 & 0.0001013 & 4.56e-05 & 0.012878 & 0.0416927 & -0.0158182 \\
\rowcolor[gray]{.8} $\epsilon$&  0.0070405 & 0.0071891 & 0.0016972 & -0.0080494 & -0.005516 & -0.0158182 & 0.1567165 \\
\hline 
\end{tabular}
\caption{Typical Covariance matrix of the parameters at the end of a run.}
\label{tab:long_cov} 
\end{table}

\begin{table}[t]
\centering
\begin{tabular}{|c | c  c  c  c  c  c  c |}
\hline 
  & $\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$ \\
\hline 
\rowcolor[gray]{.8} $\tau_0$ &   0.0189481 & -0.0014269 & -0.0011267 & -1.13e-05 & -0.0025616 & -0.0189559 & 0.0070405 \\
$\alpha$ &                       -0.0014269 & 0.0026716 & 9.93e-05 & -0.0002041 & -0.0008632 & -0.0016823 & 0.0071891 \\
\rowcolor[gray]{.8} $E_0$    &   -0.0011267 & 9.93e-05 & 0.0010701 & -0.0002277 & -0.0001177 & 0.0001013 & 0.0016972 \\
$V_0$    &                       -1.13e-05 & -0.0002041 & -0.0002277 & 0.0005401 & 4.3e-06 & 4.56e-05 & -0.0080494 \\
\rowcolor[gray]{.8} $\tau_s$ &   -0.0025616 & -0.0008632 & -0.0001177 & 4.3e-06 & 0.0128056 & 0.012878 & -0.005516 \\
$\tau_f$ &                       -0.0189559 & -0.0016823 & 0.0001013 & 4.56e-05 & 0.012878 & 0.0416927 & -0.0158182 \\
\rowcolor[gray]{.8} $\epsilon$&  0.0070405 & 0.0071891 & 0.0016972 & -0.0080494 & -0.005516 & -0.0158182 & 0.1567165 \\
\hline 
\end{tabular}
\caption{Typical Covariance matrix of the parameters at the end of a run.}
\label{tab:long_corr} 
\end{table}

To test how well the particle filter would do given much more data, and 
determine the inherent variance in the model parameters, a very long simulation
with randomly generated impulse input was created. The preprocessed timeseries
and the final estimate are shown in \autoref{fig:long_preprocessed}, 
 and the final covariance matrix of the parameters is shown in \autoref{tab:long_cov}.
Note that even though the
BOLD response converged well (\autoref{fig:long_converge}),
the parameters still have significant correlation (\autoref{tab:long_corr}). This points
to a certain amount of indeterminability of the parameters.
